\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-03 paper/ms.tex
Working on prediction section

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
the likelihood with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.

@@

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on several years of \kepler\ observations
    \citep{Pinsonneault:2014}.
    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
    regions in the marginalized planes and the histograms along the diagonal
    show the marginalized posterior for each parameter.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

\response{
This example is different from all the previous examples~--~both simulated and
real~--~because, in this case, do not set the deterministic
mean function $\mu_\bvec{\theta}$ to zero.
Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
physically significant function and the parameters are fundamental properties
of the system.
This is an example using real data~--~because we use the light curve from
\sect{rotation}~--~but we multiply these data by a simulated transiting
exoplanet model with known parameters.
This allows us to show that we can recover the true parameters of the planet
even when the transit signal is superimposed on the real variability of a
star.
}
GP modeling has been used extensively for this purpose throughout the
exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.

In \eq{gp-likelihood} the physical parameters of the exoplanet are called
$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
planet in units of the stellar radius $R_P/R_\star$, and two parameters
describing the limb-darkening profile of the star
\citep{Claret:2011,Kipping:2013}.
As in \sect{rotation}, we model the stellar variability using a GP model
with a kernel given by \eq{rot-kernel} and fit for the parameters of the
exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
simultaneously.
\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
listed in Table~\ref{tab:example-5-params} along with their priors and the
true values for $\bvec{\theta}$.}

\response{ We take a 20~day segment of the \kepler\ light curve for
KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
model with the parameters listed in Table~\ref{tab:example-5-params}.
Using these data, we maximize the joint posterior defined by the likelihood in
\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
simultaneously.
The top panel of \Figure{transit-ml} shows the data including the simulated
transit as black points with the MAP model prediction over-plotted in blue.
The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
where the MAP model has been subtracted and the MAP mean model
$\mu_\bvec{\theta}$ has been added back in.
For comparison, the transit model is over-plotted in the bottom panel of
\Figure{transit-ml} and we see no obvious correlations in the residuals.

Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
MAP parameters, we run \examplevnburn~steps of burn-in and
\examplevnsteps~steps of production MCMC using \emcee.
We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
chain and find \examplevneff~effective samples across the full chain.
\Figure{transit-corner} shows the marginalized posterior constraints on the
physical properties of the planet compared to the true values.
Even though the \celeriteterm\ representation of the stellar variability is
only an effective model, the inferred distributions for the physical
properties of the planet $\bvec{\theta}$ are consistent with the true values.
This promising result suggests that \celerite\ can be used as an effective
model for transit inference for a significant gain in computational
tractability.
Even for this example with small $N$, the cost of computing the likelihood
using \celerite\ is nearly two orders of magnitude faster than the same
computation using a general direct solver.
}

\input{figures/transit/transit-params-edited}

@@

\dfmtodo{Do this.}

@@

There are other methods of scaling GP models to large datasets and in this
section we draw comparisons between \celerite\ and other popular methods.
Scalable GP methods tend to fall into two categories: approximate and
restrictive.
\celerite\ falls into the latter category because, while the method is exact,
it requires a specific choice of stationary kernel function and it can only be
used in one-dimension.

\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}

\response{
The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
is, in practice, somewhat faster and the memory requirements are smaller.
One benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
}

\subsection{Other scalable methods}

%As discussed previously, the \celerite\ method is based upon a method
%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
%In the case where the kernel function has a single term of the form given in
%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
%and the Rybicki \& Press method.
%Their method does not require solving the extended matrix because the matrix
%inverse can be calculated analytically, but the method is made slightly more
%complicated when white noise is included.
%We implement their full method and find empirically that it is more
%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
%but we note that the requirement of a single, real, exponential kernel is
%extremely restrictive and this model cannot describe quasi-periodic
%variability.

@@

>>>>

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term in $-2\ln{\mathcal{L}}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
This last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on several years of \kepler\ observations
    \citep{Pinsonneault:2014}.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

In this example, we inject the signal of a simulated exoplanet transit into a
real \kepler\ light curve and then demonstrate that we can recover the true
physical parameters of the exoplanet while modeling the stellar variability
using \celerite.
This example is different from all the previous examples~--~both simulated and
real~--~because, in this case, we are uninterested in the parameters of the
covariance model.
Instead, we're interested in inferring constraints on the parameters of the
mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
signal caused by the transit of the planet.
We say ``mean model'' here to unambiguously distinguish it from the
``covariance model'', but this is what would generally be called the ``model''
in other analyses.
In this case, we have a light curve of a rotating star with a planet
transiting and we would like to fit for the physical parameters of the planet
while taking our uncertainty about the star's variability into account.
GP modeling has been used for this purpose throughout the exoplanet literature
\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
Foreman-Mackey:2016b, Grunblatt:2016}.

In \eq{gp-likelihood} the physical parameters of the planet are called
$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
$\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
$t_0$, an impact parameter $B$, the radius of the planet in units of the
stellar radius $R_P/R_\star$, and several parameters describing the
limb-darkening profile of the star \citep{Claret:2011}.
As in the previous example, we model the stellar variability using a GP model
with a kernel given by \eq{rot-kernel} and fit for the parameters of the
planet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
simultaneously.

We take a month-long segment of the \kepler\ light curve for
KIC~1430163~--~the target from \sect{rotation}~--~and multiply it by a
simulated transit model with known parameters.
The fact that we are using a simulated transit enables us to confirm that we
correctly measure the planet parameters but we do not use this knowledge to
inform our fit.
The top panel of \Figure{transit-ml} shows the data including the simulated
transit.
The bottom panel shows the same data with the maximum likelihood prediction
for the stellar variation subtracted.
To find this model, we maximized the likelihood function in \eq{gp-likelihood}
with respect to the transit parameters $\bvec{\theta}$ and the variability
parameters $\bvec{\alpha}$ simultaneously.
\Figure{transit-corner} shows the marginalized posterior constraints on the
physical properties of the planet compared to the true values.
This procedure produces estimates of the planet parameters that are consistent
with the true values using \celerite\ as an effective model for the stellar
variability.

\newpage

\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
and the impact parameter.}

\input{figures/transit/transit-params}

@@

\dfmtodo{The D in Table 6 may be confused with the D used in the Cholesky solver.}

@@

There are many other methods of scaling GP models to large datasets and in
this section we draw comparisons between \celerite\ and other popular methods.
Scalable GP methods tend to fall into two categories: approximate and
restrictive.
\celerite\ falls into the latter category because, while the method is exact,
it requires a specific choice of stationary kernel function and it can only be
used in one-dimension.

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form given in
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically, but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
is, in practice, somewhat faster and the memory requirements are smaller.
One benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
